---
title: "Randomization inference"
author: "David Linnard Wheeler"
date: "6/14/2021"
output: html_document
---

```{r echo=FALSE, results='hide',message=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=5) 
```


### Imagine that we want to determine if some groups are _different_ from each other.

```{r echo=FALSE, results='hide',message=FALSE}
# Invoke libraries
library("data.table")
library("ggplot2")
library("tidyverse")
library("dplyr")
library("knitr")
```

Let's load some data to demonstrate:

```{r echo=FALSE, results='hide'}
d <- fread("/Users/davidwheeler/Desktop/DataScience/BERKELEY/W241/PS_2/w241-ps2-spring2020-section5/data/list_data_2019 (1).csv")
# Rename variables
setnames(d, "uniform_price_auction", "auction_type")
```

```{r}
# The top of the data table/frame
head(d)
```
<div style="margin-bottom:100px;">

Next, let's formulate a clear question and testable hypothesis:

**Question**: Is there a difference in bids between different auction types?

**Sharp $H_0$**: $\tau_i = Y_i(0) - Y_i(1) = 0$ for each subject/unit.

- Note that this is a **much** stronger assertion than the typical, difference in means, hypothesis we generally pose.  
- Also, note that randomization inference is flexible- we can test many types of hypotheses.


```{r echo=FALSE, results='hide'}

str(d)
# Coerce x to factor
d$auction_type = factor(d$auction_type)
# Sanity check
str(d)
```
</div>


<div style="margin-bottom:100px;">
#### Before we do anything fancy let's just see if there is _a_ difference between the two treatments
```{r}
# Group mus and sigmas by auction type
knitr::kable(d[ , .(mu_hat=mean(bid),
      sigma_hat=sqrt(var(bid))),
   by = auction_type ],
   format = "html",
   table.attr = "style='width:30%;'")
```
</div>
<div style="margin-bottom:100px;">
Numerically, the groups _appear_ different. Let's look closer:
```{r fig1, fig.height = 3, fig.width = 5}
d %>%
ggplot(aes(x=auction_type, y=bid,
           group=auction_type)) +
geom_boxplot() +
  geom_jitter(size=5, pch=20, alpha=1/2,
              position=position_jitter(w=0.1, h=0.1))+
  theme_bw(base_size = 15) +
  labs(x="auction type", y="bid($)")
```
</div>
<div style="margin-bottom:100px;">
#### So, are they _really_ different
```{r}
# # 1st solution
# ate <- d[ , mean(bid), keyby = .(auction_type)][ , diff(V1)]
# ate
# # 2nd solution
# TCmeans <- d[ , .(mu = mean(bid)), keyby=.(auction_type)]
# ate <- TCmeans[ , diff(mu)]
# ate
# 3rd solution
ts <- mean(d$bid[d$auction_type == 1]) - mean(d$bid[d$auction_type == 0])
ts
```


##### One way to test for a difference-in-means with a t-test:
```{r}
# t-test 1
#d[ , t.test(bid ~ auction_type)]
# t-test 2
t <- t.test(bid ~ auction_type, data=d)

# t-test statistic
# t$statistic[[1]]

# t-test statistic
# t$conf.int[1:2]

```

This method is OK if you satisfy the requisite assumptions and sometimes even if you do not...

But these assumptions can be restrictive.

This is where randomization inference can come to the rescue.
</div>
<div style="margin-bottom:300px;">

##### Another solution is to use randomization inference (RI).

There are many ways to introduce RI. Most are intuitive.

First, let's go back in time to just before treatment assignment. 

- Note that there is >1 way to assign treatments to units/subjects.

- More formally, since there are `r sum(d$auction_type==1)` units each in treatment and control and `r nrow(d)` total observations, there are $68\choose 34$ = $\frac{64!}{34!34!}$ = `r choose(68, 34)` possible treatment assignments!

- That is a lot of ways to assign treatments to units. We just _used_ one way. 

- RI takes advantage of these possible but unobserved counterfactual treatment assignments. It asks "how likely are we to observe the treatment effect we detected ($test\;statistic$ = `r ts`) in all of the experiments we could have conducted with different treatment assignments."

</div>
<div style="margin-bottom:300px;">

- To answer this question, we need 3 things:

  - a test statistic
  - a null hypothesis
  - a measure of extremeness


- Let's go through each one in turn.

</div>
<div style="margin-bottom:300px;">

- Test statistic
  - This could be anything like, say a difference-in-means:

```{r}
# 1 solution
ts <- mean(d$bid[d$auction_type == 1]) - mean(d$bid[d$auction_type == 0])
# 2 Solution
tslm <- lm(bid ~ auction_type, data=d)
# all together now
knitr::kable(data.table(method = c("ByHand","LinearModel"),
           estimate = c(ts, tslm$coefficients[[2]])),
   format = "html",
   table.attr = "style='width:30%;'")
```
</div>
<div style="margin-bottom:300px;">

- Null distribution

  - This is the _meat and potatoes_ of RI. 
  - With other methods, like a t-test, we compute our test statistic,$t$, find it under the sampling distribution of the statistic (a $t$ distribution) and use the probability of observing our $t$ or one more extreme under the null as evidence against our $H_0$.
  - With RI, we use a simpler approach. 
    - First, we assume the sharp null ($\tau_i = Y_i(0) - Y_i(1) = 0$ for each subject/unit) is true. 
    - If the sharp null is true than then treatment has no effect on each unit. Thus, we can shuffle (ahem, _permute_) the treatment vector, calculate our test statistic after each permutation, and repeat _many_ times. This procedure yields the sampling distribution of our test-statistic. This DIY null distribution describes all of the possible test-statistics that could have arisen with _our_ data.
    - Next we just find our original test-statistic (the one we calculated before the permutations, here its `r ts`).
    
</div>
<div style="margin-bottom:300px;">

- This is how this works:
```{r}
 
# example 1
# Null distribution
null_dist1 <- NA 
# For each randomization, r, from 1 to 10000
for(r in 1:100000) { 
  # populate this empty vector, ri_distribution, with ates after each shuffle
  null_dist1[r] <- d[ , .(group_mean = mean(bid)), 
    keyby = .(sample(auction_type))][ , diff(group_mean)]
}
```

```{r}

# example 2
# Randomization inference
# Vector  of NAs to replace with estimates under randomization
null_dist2 <- rep(NA, 10000)

# For each test statistic, ATE, from 1 to 10,000
for(a in 1:10000) {
  # Assign to treatment, a random vector of 0s and 1s.
  # Let the number of 0s and 1s be equal to the number of winners (1) and non-winners (0)
  treatment <- sample(c(rep(0, sum(d$auction_type == 0)),
                        rep(1, sum(d$auction_type == 1))),
                  replace = TRUE)
  # Assign ate to the difference in mean views for winners and non-winners
  null_dist2[a] <- d[treatment==1, .(mean(bid))] - d[treatment==0, .(mean(bid))]
}
```

```{r}

# Example 3
# Build vector of randomized treatment assignments
sim.study <- function() {
  # data table with two columns
                      # vector of RANDOMIZED treatment assignments
  d.sim <- data.table(auction_type = sample(c(rep(0,length(d$auction_type)/2),
                                          rep(1,length(d$auction_type)/2))),
                      # vector of ORIGINAL response/label/dependent variable
                      bid = d$bid)
  # Group means
  group_means <- d.sim[ , .(mean_bids = mean(bid)),
                keyby = auction_type]# [, diff(group_mean)]
  # test statistics
  tss <- group_means[auction_type == 1]$mean_bids - group_means[auction_type == 0]$mean_bids
  # Vector of ATEs
  return(tss)
  
}

# Run simulations
null_dist3 <- t(replicate(10000, sim.study()))
```




Let's visualize our DIY null distribution:
```{r fig2, fig.height = 3, fig.width = 5}

#par(mfrow=c(3,1))
# All ATEs
hist(null_dist1,
     xlim=c(-15,15), main=" ")
# Our original ATE
abline(v=ts, col="red")

# # All ATEs
# hist(as.numeric(null_dist2), 
#      xlim=c(-15,15), main=" ")
# # Our original ATE
# abline(v=ts, col="red")
# 
# # All ATEs
# hist(as.numeric(null_dist3[1,]), 
#      xlim=c(-15,15), main=" ")
# # Our original ATE
# abline(v=ts, col="red")
```

<div style="margin-bottom:300px;">

#### How do we interpret this?

- This is where the measure of extremeness comes in to play.
  - In general we have two choices:
    1. Interpret large OR small simulated test statistics as _directional_ evidence against the null
    2. Interpret large AND small simulated test statistics as _categorical_ evidence against the null
  

To do this, we can just count the number of simulated test statistics that are larger in absolute value than our actual test statistic.
```{r}
# number of larger ates
larger_ts1 <- sum(null_dist1 > abs(ts))
larger_ts1

# number of larger ates
larger_ts2 <- sum(abs(null_dist1) > abs(ts))
larger_ts2
```

</div>
<div style="margin-bottom:300px;">

Further we can calculate confidence intervals
```{r}

# CI
ri.CI <- quantile(null_dist1,
                  probs = c(0.025, 0.975))
# Centered about the ate
ri.CI <- ri.CI + ts

ri.CI[[1]]; ri.CI[[2]]
```

</div>
<div style="margin-bottom:300px;">

And p-values
```{r}
# 1-tailed P-value
p_value_1tail <-  sum(null_dist1 >= abs(ts))/10000 # mean(ri_distribution >= abs(ate))

# 2-tailed P-value
p_value_2tailed <- sum(abs(null_dist1) >= abs(ts))/10000 # mean(abs(ri_distribution) >= abs(ate))

# Compare
knitr::kable(data.table("Test" = c("T.test","RI.1tail", "RI.2tailed"),
            "statistic" = c(t$statistic[[1]], ts, ts),
           "confidence_interval_UB" = c(t$conf.int[1], ri.CI[[2]], ri.CI[[2]]),
            "confidence_interval_UB" = c(t$conf.int[2], ri.CI[[1]], ri.CI[[1]]),
            "p-values" = c(p_value_1tail, p_value_2tailed, t$p.value)),
   format = "html",
   table.attr = "style='width:70%;'")

```

</div>
<div style="margin-bottom:300px;">

Does this make sense? 

What did we learn?

Three benefits of RI:
- Its flexible.
- Its agnostic to many assumptions.
- It is exact because the set of all possible random assignments fully describe the sampling distribution under the null hypothesis. Many other methods derive the shape of sampling distributions by approximation.

</div>
<div style="margin-bottom:300px;">

#### Resources:
- [Fisher’s Randomization Inference](https://www.mattblackwell.org/files/teaching/s05-fisher.pdf)
- [R package](http://alexandercoppock.com/ri2/articles/ri2_vignette.html)
- [Field experiments: design, analysis, and interpretation](https://www.amazon.com/Field-Experiments-Design-Analysis-Interpretation/dp/0393979954)
